Efficient spherical Convolutional Neural Networks with Healpix sampling for cosmological applications

### 2018-04-25 scnn.tex
starting section graph

The first approach, leveraging the continuous spherical convolution, has been
proposed in~\cite{cohen2017convolutional,cohen2018spherical}. These
contributions propose to use a convolution that is rotational equivariant on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The convolution is performed by a spherical Fourier transform (i.e. a projection
on the spherical harmonics), a  multiplication in the spectral domain and an
inverse spherical Fourier tranform. Hence the main computational cost of a
convolution comes from the two Fourier transforms. Fortunately, the Fourier
transform can be accelerated for special samplings set (theoretically including
HealPix). The main advantage of this approach is that it provides a
mathematically well defined rotational equivariant network. Nevertheless, even
with these Fourier transform acceleration, the convolutions remain expensive in
comparison to a traditional 2D convolution, limiting the practical use of this
approach. While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible.

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2D convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2D convolution, it is very likely to be the
most efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere. \nati{Michael: Under some very
specific hypothesis, this second approach is a particular case of our method,
i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
on a grid == 2d convolution. Do you think we should mention that? I think we
should not.}

@@

As presented in~\cite{cohen2018spherical}, the most mathematical approach to extend the convolution on a sphere is to use a spherical Fourier transform. The convolution is then simply defined as the product in the spectral domain. This approach requires one Fourier and one inverse Fourier transform per convolution which is, even with accelerated algorithms, rather expensive. For 2-dimensional images, an efficient convolution can be achieved when the convolution kernel is localized (for example a 5x5 pixel patch) by doing the computation directly in the signal domain. Unfortunately, this approach cannot be directly extended to the spherical case. Hence, the main idea of this contribution is to leverage graph signal processing~\cite{shuman2013emerging} to define a spherical convolution that can be computed directly in the signal domain.

In this section, we follow classical constructions presented in~\cite{cohen2018spherical,...} and \nati{show} that the resulting spherical convolution is close to one corresponding to spherical approach.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\includegraphics[width=0.45\textwidth]{figures/half_graph_4.pdf}
\caption{Left: Healpix graph for the full sphere using a nside parameter of $4$. Right: Top view of a Healpix graph for half the sphere using a nside parameter of $4$.}
\label{fig:healpix_graph_4}
\end{figure}

From the ubiquitous HealPix sampling~\citep{gorski2005healpix}, we create a weighted undirected graph where each pixel is a node (vertex) that is connected to his $8$ or $7$ closest neighbors.\footnote{For some pixels, the $8^{th}$ nearest neighbor is not well defined.} Given the set of nearest neighbors, we define the weight matrix $W$ using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$ on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected pixels $i$ and $j$. The degree of a node (or a pixel) is defined as $d_i = \sum_j W[i,j]$. The degree matrix $D$ is the diagonal matrix where $D_{ii}=d_i$.

Figure Graph (show also the triangle, separate plot?) + sampling

>>>>

The first approach, leveraging the continuous spherical convolution, has been
proposed in~\cite{cohen2017convolutional,cohen2018spherical}. These
contributions propose to use a convolution that is rotational equivariant on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The convolution is performed by a spherical Fourier transform (i.e. a projection
on the spherical harmonics), a  multiplication in the spectral domain and an
inverse spherical Fourier tranform. Hence the main computational cost of a
convolution comes from the two Fourier transforms. Fortunately, the Fourier
transform can be accelerated for special samplings set (theoretically including
HealPix). The main advantage of this approach is that it provides a
mathematically well defined rotational equivariant network. Nevertheless, even
with these Fourier transform acceleration, the convolutions remain expensive in
comparison to a traditional 2-dimensional convolution, limiting the practical use of this
approach. While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible.

A second direction has been followed in~\cite{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid and the “cubed
sphere” grid defined by Ronchi et. al~\cite{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere. \nati{Michael: Under some very
specific hypothesis, this second approach is a particular case of our method,
i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
on a grid == 2d convolution. Do you think we should mention that? I think we
should not.}

@@

As presented in~\cite{cohen2018spherical}, the most mathematical approach to
extend the convolution on a sphere is to use a spherical Fourier transform. The
convolution is then simply defined as the product in the spectral domain. This
approach requires one Fourier and one inverse Fourier transform per convolution
which remain, even with accelerated algorithms, expensive. For 2-dimensional
images, an efficient convolution can be achieved when the convolution kernel is
localized (for example a 5x5 pixel patch) by doing the computation directly in
the signal domain. Unfortunately, this approach cannot be directly extended to
the spherical case. Hence, the main idea of this contribution is to leverage
graph signal processing~\cite{shuman2013emerging} to define a spherical
convolution that can be computed directly in the signal domain.

@@

In the classical 2-dimenstional convolution, each pixel is connected with the
same weight to its 4 closest neighbors. We use a graph to generalize this to the
irregular sphear sampling. From the ubiquitous HealPix
sampling~\citep{gorski2005healpix}, we create a weighted undirected graph where
each pixel is a node (vertex) that is connected to his $8$ or $7$ closest
neighbors.\footnote{For some pixels, the $8^{th}$ nearest neighbor is not well
defined.} Given the set of nearest neighbors, we define the weight matrix $W$
using the following scheme
\begin{equation}
W[i,j]=\begin{cases}
e^{-\frac{\|x_i-x_j\|_2^2}{\sigma^2}} & \text{if pixels $i$ and $j$ are connected, and}\\
0 & \text{otherwise.}\\
\end{cases}
\end{equation}
Here $x_i$ is a 3-dimensional vector encoding the coordinate of the pixels $i$
on the sphere and $\sigma$ is the mean of $\|x_i-x_j\|_2$ over all connected
pixels $i$ and $j$. The degree of a node (or a pixel) is defined as $d_i =
\sum_j W[i,j]$. The degree matrix $D$ is the diagonal matrix where $D_{ii}=d_i$. 
An example of the resulting graph is displayed in Figure~\ref{fig:healpix_graph_4}.

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\includegraphics[width=0.45\textwidth]{figures/half_graph_4.pdf}
\caption{
% Left: Healpix graph for the full sphere using a nside parameter of $4$. 
Top view of a Healpix graph for half the sphere using a nside parameter of $4$.}
\label{fig:healpix_graph_4}
\end{figure}
